The CrypTool Script: \\ Cryptography, Mathematics and More \\ \vspace{30pt} {\large Background reading\\%*[-8pt] for the free e-learning program CrypTool\\ (10th edition -- distributed with CrypTool version 1.4.30)} \vspace{60pt}

### 2009-08-14 script/en/chapters/numbertheory.tex
fixed math mode in headings resulting in broken PDF bookmarks with \texorpdfstring; fixed float warnings by replacing placement [h] with [ht]; fixed warning about PDF version 1.5 by changing it to 1.4 in coverv1_*.pdf
CT script: Sage parts finished in (en)

\begin{table}[h]
\begin{center}
\begin{tabular}{|l||@{\:}c@{\:}|@{\:}c@{\:}|@{\:}c@{\:}|@{\:}c@{\:}|@{\:}c@{\:}|@{\:}c@{\:}|@{\:}c@{\:}|@{\:}c@{\:}|@{\:}c@{\:}|@{\:}c@{\:}|@{\:}c@{\:}|@{\:}c@{\:}|@{\:}c@{\:}|@{\:}c@{\:}|@{\:}c@{\:}|@{\:}c@{\:}||@{\:}c@{\:}|@{\:}c@{\:}|}
\hline 
i                   & 1  & 2  & 3  & 4  & 5  & 6  & 7  & 8  & 9 & 10 & 11 & 12 & 13 & 14 & 15 & 16  & 17 & 18 \\
\hline
\hline  
$5*i$                & 5 & 10 & 15 & 20 & 25 & 30 & 35 & 40 & 45 & 50 & 55 & 60 & 65 & 70 & 75 & 80  & 85 & 90   \\
remainder                & 5 & 10 & 15  & 3  & 8 & 13  & {\bf 1}  & 6 & 11 & 16  & 4  & 9 & 14  & 2  & 7 & 12 & 0  & 5   \\
\hline
$6*i$                 & 6 & 12 & 18 & 24 & 30 & 36 & 42 & 48 & 54 & 60 & 66 & 72 & 78 & 84 & 90 & 96 & 102 & 108   \\
remainder                & 6 & 12  & {\bf 1}  & 7 & 13  & 2  & 8 & 14  & 3  & 9 & 15  & 4 & 10 & 16  & 5 & 11 & 0  & 6   \\
\hline
\end{tabular}
\end{center} 
\caption{Multiplication table modulo $17$ (for $a=5$ and $a=6$)}\label{SrcArith1a}\label{mulmod17}
\end{table}
Between $i=1, \cdots, m$, all values between $0, \cdots, m-1$ occur for the remainders, because both $5$ and $6$ are 
also relatively prime\index{Relatively prime} to the modulus $m=17$.
\enlargethispage{0.5cm}

@@

\begin{table}[h]
\begin{center}                                                                          
\begin{tabular}{|l||@{\:}c@{\:}|@{\:}c@{\:}|@{\:}c@{\:}|@{\:}c@{\:}|@{\:}c@{\:}|@{\:}c@{\:}|@{\:}c@{\:}|@{\:}c@{\:}|@{\:}c@{\:}|@{\:}c@{\:}|@{\:}c@{\:}|@{\:}c@{\:}||@{\:}c@{\:}|@{\:}c@{\:}|@{\:}c@{\:}|@{\:}c@{\:}|@{\:}c@{\:}|@{\:}c@{\:}|}
\hline 
i                    & 1  & 2  & 3  & 4  & 5  & 6  & 7  & 8  & 9 & 10 & 11 & 12 & 13 & 14 & 15 & 16  & 17  & 18 \\
\hline 
\hline 
$5*i$                  & 5 & 10 & 15 & 20 & 25 & 30 & 35 & 40 & 45 & 50 & 55 & 60 & 65 & 70 & 75 & 80 & 85  & 90 \\
remainder                 & 5 & 10  & 2  & 7  & 12  & 4 & 9  & {\bf 1}  & 6  & 11 & 3  & 8  & 0 & 5  & 10  & 2   & 7   & 12 \\
\hline 
$6*i$                  & 6 & 12 & 18 & 24 & 30 & 36 & 42 & 48 & 54 & 60 & 66 & 72 & 78 & 84 & 90 & 96 & 102 & 108 \\
remainder                 & 6  & 12  & 5  & 11  & 4  & 10  & 3  & 9  & 2  & 8  & {\bf 1}  & 7  & 0  & 6  & 12  & 5   & 11   & 4 \\
\hline 
\end{tabular}
\end{center} 
\caption{Multiplication table modulo $13$ (for $a=5$ and $a=6$)}\label{SrcArith1b}\label{mulmod13}
\end{table}

@@

\begin{table}[h]
\begin{center}                                                                          
\begin{tabular}{|l||@{\:}c@{\:}|@{\:}c@{\:}|@{\:}c@{\:}|@{\:}c@{\:}|@{\:}c@{\:}|@{\:}c@{\:}|@{\:}c@{\:}|@{\:}c@{\:}|@{\:}c@{\:}|@{\:}c@{\:}|@{\:}c@{\:}||@{\:}c@{\:}|@{\:}c@{\:}|@{\:}c@{\:}|@{\:}c@{\:}|@{\:}c@{\:}|@{\:}c@{\:}|@{\:}c@{\:}|}
\hline 
i                    & 1  & 2  & 3  & 4  & 5  & 6  & 7  & 8  & 9 & 10 & 11 & 12 & 13 & 14 & 15 & 16  & 17 & 18 \\
\hline 
\hline 
$5*i$                  & 5 & 10 & 15 & 20 & 25 & 30 & 35 & 40 & 45 & 50 & 55 & 60 & 65 & 70 & 75 & 80 & 85  & 90 \\
remainder                 & 5 & 10  & 3  & 8  & {\bf 1}  & 6 & 11  & 4  & 9  & 2  & 7  & 0  & 5 & 10  & 3 & 8   & 1   & 6 \\
\hline 
$6*i$                  & 6 & 12 & 18 & 24 & 30 & 36 & 42 & 48 & 54 & 60 & 66 & 72 & 78 & 84 & 90 & 96 & 102 & 108 \\
remainder                 & 6  & 0  & 6  & 0  & 6  & 0  & 6  & 0  & 6  & 0  & 6  & 0  & 6  & 0  & 6  & 0   & 6   & 0 \\
\hline 
\end{tabular}
\end{center} 
\caption{Multiplication table modulo $12$ (for $a=5$ and $a=6$)}
\label{mulmod12}
\end{table}
We have calculated $(5 * i)$ (mod $12$)  and  $(6*i)$ (mod $12$).
Between $i=1, \cdots, m$, not all values between $0, \cdots, m-1$ occur and $6$ does not have an inverse mod $12$, 
because $6$ and the modulus $m=12$ are not co-prime\index{Number!co-prime}.

@@

% ++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++
\section{Groups and modular arithmetic in $\mathbb{Z}_n$ and $\mathbb{Z}_n^*$}
\index{Group}
Mathematical ``{\em groups}'' play a decisive role in number theory and cryptography. We only talk of groups if, for a 
defined set and a defined relation (an operation such as addition or multiplication), the following properties are fulfilled:

@@

\begin{table}[h]
\begin{center}
\begin{tabular}{|l||c|c|c|c|c|c|c|c|c|c|c|c|c|c|}
\hline
              & i=1 & i=2 & i=3 & i=4 & i=5 & i=6 & i=7 & i=8 & i=9 & i=10  & $ord_{11}(a)$\\
\hline
\hline
a=1           & {\bf 1}  & 1    & 1  & 1    & 1    & 1    & 1  & 1    & 1  & 1     & 1   \\
\hline
a=2           & 2  & 4    & 8  & 5   & 10    & 9    & 7  & 3    & 6  & {\bf 1}    & 10  \\
\hline
a=3           & 3  & 9    & 5  & 4 & {\bf 1} & 3    & 9  & 5    & 4  & 1     & 5   \\
\hline
a=4           & 4  & 5    & 9  & 3 & {\bf 1} & 4    & 5  & 9    & 3  & 1    & 5 \\
\hline
a=5           & 5  & 3    & 4  & 9 & {\bf 1} & 5    & 3  & 4    & 9  & 1    & 5   \\
\hline
a=6           & 6  & 3    & 7  & 9   & 10    & 5    & 8  & 4    & 2  & {\bf 1}    & 10  \\
\hline
a=7           & 7  & 5    & 2  & 3   & 10    & 4    & 6  & 9    & 8  & {\bf 1}    & 10  \\
\hline
a=8           & 8  & 9    & 6  & 4   & 10    & 3    & 2  & 5    & 7  & {\bf 1}    & 10  \\
\hline
a=9           & 9  & 4    & 3  & 5 & {\bf 1} & 9    & 4  & 3    & 5  & 1    & 5   \\
\hline
a=10         & 10  & {\bf 1}   & 10  & 1   & 10    & 1   & 10  & 1   & 10  & 1    & 2   \\
\hline
\end{tabular}
\end{center}
\caption{Values of $a^i {\rm ~mod~} 11,  1 \leq a,i<11$ and according order of $a$ mod $m$}
\label{SrcArith3a}\label{expmod11}
\end{table}

@@

\begin{table}[h]
\begin{center}
\begin{tabular}{|l||c|c|c|c|c|c|c|c|c|c|c|c|c|c|c|c|c|c|c|c|c|c|c|c|c|}
\hline
 $a\setminus i$ & 1            & 2            & 3 & 4 & 5 & 6 & 7 & 8 & 9 & 10 & 11 & 12     & $ord_{45}(a)$       & $J(45)$ \\
\hline
\hline                                                       
1             & 1              & 1   & 1   & 1   & 1   & 1   & 1   & 1   & 1    & 1    & 1    & 1 & 1              & 24  \\
\hline
2             & 2              & 4   & 8  & 16  & 32  & 19  & 38  & 31  & 17   & 34   & 23    & 1 & 12             & 24 \\
\hline
3             & 3              & 9  & 27  & 36  & 18   & 9  & 27  & 36  & 18    & 9   & 27   & 36  & ---            & 24 \\
\hline
4             & 4             & 16  & 19  & 31  & 34   & 1   & 4  & 16  & 19   & 31   & 34    & 1  & 6              & 24 \\
\hline
5             & 5             & 25  & 35  & 40  & 20  & 10   & 5  & 25  & 35   & 40   & 20   & 10  & ---            & 24 \\
\hline
6             & 6             & 36  & 36  & 36  & 36  & 36  & 36  & 36  & 36   & 36   & 36   & 36  & ---            & 24 \\
\hline
7             & 7              & 4  & 28  & 16  & 22  & 19  & 43  & 31  & 37   & 34   & 13    & 1  & 12             & 24 \\
\hline
8             & 8             & 19  & 17   & 1   & 8  & 19  & 17   & 1   & 8   & 19   & 17    & 1  & 4              & 24 \\
\hline
9             & 9             & 36   & 9  & 36   & 9  & 36   & 9  & 36   & 9   & 36    & 9   & 36  & ---            & 24 \\
\hline
10           & 10             & 10  & 10  & 10  & 10  & 10  & 10  & 10  & 10   & 10   & 10   & 10  & ---            & 24 \\
\hline
11           & 11             & 31  & 26  & 16  & 41   & 1  & 11  & 31  & 26   & 16   & 41    & 1  & 6              & 24 \\
\hline
12           & 12              & 9  & 18  & 36  & 27   & 9  & 18  & 36  & 27    & 9   & 18   & 36  & ---            & 24 \\
\hline
\end{tabular}
\end{center}
\caption{Values of $a^i{\rm ~mod~}45, 1\leq a,i<13$}
\label{SrcArith3b}\label{expmod45}
\end{table}

@@

\item[{\bf 2.}] Select an arbitrary $e \in \{2, \cdots, n-1\}$ such
                that\footnote{%
                \label{foot:Selection-of-e}%
                It is recommended by cryptanalytic reasons, but not
                necessary to make RSA work, to select $e$ such that: \\
                $\max(p,q)  <  e <  J(n) - 1$.}: \\
                $e$ is relatively prime to $J(n) = (p-1)*(q-1)$. \\
                We can then ``throw away'' $p$ and $q$.\footnote{%
                The procedure also allows us to select $d$ freely and then 
                calculate $e$. However, this has practical disadvantages.
                We usually want to be able to encrypt messages ``quickly'',
                which is why we choose a public exponent $e$ such that 
                it has a short bit length compared to the modulus $n$ and 
                as few binary ones as possible (e.g. $2^{16} + 1$).
                So a fast exponentiation is possible when encrypting.                
                We want to select the publicly known $e$ to be an
                advantageous value that allows the exponential calculation
                to be performed quickly during encryption. The prime numbers
                $3, 17$ and $65537$ have proved to be particularly practical
                for this purpose.
                The most often used number is $65537 = 2^{16}+1$, or in
                binary: 
                $10\cdots 0\cdots 01$ (this number is prime and therefore
                relatively prime to many other numbers).
                }
\item[{\bf 3.}] Select $d \in \{1, \cdots, n-1\}$ with $e*d \equiv 1  
                {\rm ~(mod~} J(n))$, i.e. $d$ is the multiplicative inverse
                of $e$ modulo $J(n)$\footnote{For reasons of security, 
                $d$ should not be too small.}$^,$\footnote{We start by determining either $d$ or $e$ depending 
                on the implementation.}. We can then ``throw away'' $J(n)$.
\begin{itemize}
    \item[] $\rightarrow (n, e)$ is the public key $P$.
        \item[] $\rightarrow (n, d)$ is the secret key $S$ (only $d$ must be kept secret).
\end{itemize}
\item[{\bf 4.}] For encryption, the message represented as a (binary) number is divided into parts such that each part of 
                the number is less than $n$.
\item[{\bf 5.}] Encryption of the plaintext (or the parts of it) $M \in \{1, \cdots, n-1\}$:
                $$C = E ( (n, e); M ) := M^e {\rm ~(mod~} n).$$
\item[{\bf 6.}] For decryption, the cipher text represented as a binary number is divided into parts such that each part 
                of the number is less than $n$.
\item[{\bf 7.}] Decryption of the cipher text (or the parts of it) $C \in \{1, \cdots, n-1\}$:
                $$M = D ( (n, d); C ) := C^d {\rm ~(mod~} n).$$
\end{itemize}
The numbers $d, e$ and $n$ are usually extremely large (e.~g.\ $d$ and $e$ $300$ bits, $n$ $600$ bits).

@@

\noindent Concretely for the RSA method does this mean:
\vskip - 1em
\begin{enumerate}
\item it is hard to calculate $J(n)$ for big compounds $n$ and
\item it is hard to calculate the prime factors of big compounds $n$ (Factorization!factorization problem).
\end{enumerate}
\vskip +1em

@@

\begin{table}[h]
\begin{center}
\begin{tabular}{|c|cccc|}
\hline
	& {\bf Decimal digits} & {\bf Binary digits} & {\bf Factored on} & {\bf Factored by} \\
\hline
	C307 & 		307 & 1017 & May, 2007 & Jens Franke et al. \\
	RSA-200 & 	200 & 663 & May, 2005 & Jens Franke et al. \\
	RSA640\footnotemark & 	193 & 600 & November, 2005 & Jens Franke et al. \\
	C176 & 		176 & 583 & May, 2005 & Kazumaro Aoki et al. \\
	RSA576 & 	174 & 576 & December, 2003 & Jens Franke et al. \\
	RSA-160 & 	160 & 530 & April, 2003 & Jens Franke et al. \\
	C158 & 		158 & 523 & January, 2002 & Jens Franke et al. \\
	RSA-155	&	155 & 512 & August, 1999 & Herman te Riele et al. \\
\hline
\end{tabular}
\caption{The current factoring records (as of July 2009)}    % Eyecatcher_neue-Mersenne 
\label{factorizationrecords}
\end{center}
\end{table} 
\footnotetext{%
A research group of the GISA solved this challenge which was awarded with
20,000 US dollar using the GNFS method. The researchers needed about five
months to divide this number into its both 320 bit long prime factors.\\
RSA Labs offers its challenges since the beginning of the 1990th.
The next bigger challenge awarded with 30,000 US dollars is RSA-704.\\
See:
\href{http://www.heise.de/newsticker/meldung/print/65957}
{\texttt{http://www.heise.de/newsticker/meldung/print/65957}}
}

@@

\begin{table}[h]
\begin{center}
\begin{tabular}{|c|l||c|l|}
\hline
Character & Numerical value & Character & Numerical value \\
\hline
\hline
Blank    & 0   & M  & 13 \\
A        & 1   & N    & 14 \\ 
B        & 2   & O    & 15 \\ 
C        & 3   & P    & 16 \\  
D        & 4   & Q    & 17 \\ 
E        & 5   & R    & 18 \\ 
F        & 6   & S    & 19 \\  
G        & 7   & T    & 20 \\  
H        & 8   & U    & 21 \\ 
I        & 9   & V    & 22 \\   
J       & 10   & W    & 23 \\  
K       & 11   & X    & 24 \\ 
L       & 12   & Y    & 25 \\
&              & Z    & 26 \\
\hline
\end{tabular}
\end{center}
\hypertarget{Grossbuchstaben-Alphabet}{}    
\caption{Capital letters alphabet}\index{Capital letters alphabet}
\label{alphacode}
\end{table}
\vskip +20 pt

@@

% ++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++
% \vskip +12pt
\newpage
\hypertarget{NumberTheory_Appendix_F}{}  %\hypertarget{AppendixListAndDef}{}
\section{Appendix: List of the definitions and theorems formulated in this chapter}
\label{l:NumberTheory_Appendix_F}{}  %\label{l:AppendixListAndDef}{}

>>>>

\begin{table}[ht]
\begin{center}
\begin{tabular}{|l||@{\:}c@{\:}|@{\:}c@{\:}|@{\:}c@{\:}|@{\:}c@{\:}|@{\:}c@{\:}|@{\:}c@{\:}|@{\:}c@{\:}|@{\:}c@{\:}|@{\:}c@{\:}|@{\:}c@{\:}|@{\:}c@{\:}|@{\:}c@{\:}|@{\:}c@{\:}|@{\:}c@{\:}|@{\:}c@{\:}|@{\:}c@{\:}||@{\:}c@{\:}|@{\:}c@{\:}|}
\hline 
i                   & 1  & 2  & 3  & 4  & 5  & 6  & 7  & 8  & 9 & 10 & 11 & 12 & 13 & 14 & 15 & 16  & 17 & 18 \\
\hline
\hline  
$5*i$                & 5 & 10 & 15 & 20 & 25 & 30 & 35 & 40 & 45 & 50 & 55 & 60 & 65 & 70 & 75 & 80  & 85 & 90   \\
remainder                & 5 & 10 & 15  & 3  & 8 & 13  & {\bf 1}  & 6 & 11 & 16  & 4  & 9 & 14  & 2  & 7 & 12 & 0  & 5   \\
\hline
$6*i$                 & 6 & 12 & 18 & 24 & 30 & 36 & 42 & 48 & 54 & 60 & 66 & 72 & 78 & 84 & 90 & 96 & 102 & 108   \\
remainder                & 6 & 12  & {\bf 1}  & 7 & 13  & 2  & 8 & 14  & 3  & 9 & 15  & 4 & 10 & 16  & 5 & 11 & 0  & 6   \\
\hline
\end{tabular}
\end{center} 
\caption{Multiplication table modulo $17$ (for $a=5$ and $a=6$)}\label{SrcArith1a}\label{mulmod17}
\end{table}
Between $i=1, \cdots, m$, all values between $0, \cdots, m-1$ occur for the remainders, because both $5$ and $6$ are 
also relatively prime\index{Relatively prime} to the modulus $m=17$.
\enlargethispage{0.5cm}

@@

\begin{table}[ht]
\begin{center}                                                                          
\begin{tabular}{|l||@{\:}c@{\:}|@{\:}c@{\:}|@{\:}c@{\:}|@{\:}c@{\:}|@{\:}c@{\:}|@{\:}c@{\:}|@{\:}c@{\:}|@{\:}c@{\:}|@{\:}c@{\:}|@{\:}c@{\:}|@{\:}c@{\:}|@{\:}c@{\:}||@{\:}c@{\:}|@{\:}c@{\:}|@{\:}c@{\:}|@{\:}c@{\:}|@{\:}c@{\:}|@{\:}c@{\:}|}
\hline 
i                    & 1  & 2  & 3  & 4  & 5  & 6  & 7  & 8  & 9 & 10 & 11 & 12 & 13 & 14 & 15 & 16  & 17  & 18 \\
\hline 
\hline 
$5*i$                  & 5 & 10 & 15 & 20 & 25 & 30 & 35 & 40 & 45 & 50 & 55 & 60 & 65 & 70 & 75 & 80 & 85  & 90 \\
remainder                 & 5 & 10  & 2  & 7  & 12  & 4 & 9  & {\bf 1}  & 6  & 11 & 3  & 8  & 0 & 5  & 10  & 2   & 7   & 12 \\
\hline 
$6*i$                  & 6 & 12 & 18 & 24 & 30 & 36 & 42 & 48 & 54 & 60 & 66 & 72 & 78 & 84 & 90 & 96 & 102 & 108 \\
remainder                 & 6  & 12  & 5  & 11  & 4  & 10  & 3  & 9  & 2  & 8  & {\bf 1}  & 7  & 0  & 6  & 12  & 5   & 11   & 4 \\
\hline 
\end{tabular}
\end{center} 
\caption{Multiplication table modulo $13$ (for $a=5$ and $a=6$)}\label{SrcArith1b}\label{mulmod13}
\end{table}

@@

\begin{table}[ht]
\begin{center}                                                                          
\begin{tabular}{|l||@{\:}c@{\:}|@{\:}c@{\:}|@{\:}c@{\:}|@{\:}c@{\:}|@{\:}c@{\:}|@{\:}c@{\:}|@{\:}c@{\:}|@{\:}c@{\:}|@{\:}c@{\:}|@{\:}c@{\:}|@{\:}c@{\:}||@{\:}c@{\:}|@{\:}c@{\:}|@{\:}c@{\:}|@{\:}c@{\:}|@{\:}c@{\:}|@{\:}c@{\:}|@{\:}c@{\:}|}
\hline 
i                    & 1  & 2  & 3  & 4  & 5  & 6  & 7  & 8  & 9 & 10 & 11 & 12 & 13 & 14 & 15 & 16  & 17 & 18 \\
\hline 
\hline 
$5*i$                  & 5 & 10 & 15 & 20 & 25 & 30 & 35 & 40 & 45 & 50 & 55 & 60 & 65 & 70 & 75 & 80 & 85  & 90 \\
remainder                 & 5 & 10  & 3  & 8  & {\bf 1}  & 6 & 11  & 4  & 9  & 2  & 7  & 0  & 5 & 10  & 3 & 8   & 1   & 6 \\
\hline 
$6*i$                  & 6 & 12 & 18 & 24 & 30 & 36 & 42 & 48 & 54 & 60 & 66 & 72 & 78 & 84 & 90 & 96 & 102 & 108 \\
remainder                 & 6  & 0  & 6  & 0  & 6  & 0  & 6  & 0  & 6  & 0  & 6  & 0  & 6  & 0  & 6  & 0   & 6   & 0 \\
\hline 
\end{tabular}
\end{center} 
\caption{Multiplication table modulo $12$ (for $a=5$ and $a=6$)}
\label{mulmod12}
\end{table}
We have calculated $(5 * i)$ (mod $12$)  and  $(6*i)$ (mod $12$).
Between $i=1, \cdots, m$, not all values between $0, \cdots, m-1$ occur and $6$ does not have an inverse mod $12$, 
because $6$ and the modulus $m=12$ are not co-prime\index{Number!co-prime}.

@@

% ++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++
\section{Groups and modular arithmetic in \texorpdfstring{$\mathbb{Z}_n$ and $\mathbb{Z}_n^*$}
                                                         {Zn and Zn*}}
\index{Group}
Mathematical ``{\em groups}'' play a decisive role in number theory and cryptography. We only talk of groups if, for a 
defined set and a defined relation (an operation such as addition or multiplication), the following properties are fulfilled:

@@

\begin{table}[ht]
\begin{center}
\begin{tabular}{|l||c|c|c|c|c|c|c|c|c|c|c|c|c|c|}
\hline
              & i=1 & i=2 & i=3 & i=4 & i=5 & i=6 & i=7 & i=8 & i=9 & i=10  & $ord_{11}(a)$\\
\hline
\hline
a=1           & {\bf 1}  & 1    & 1  & 1    & 1    & 1    & 1  & 1    & 1  & 1     & 1   \\
\hline
a=2           & 2  & 4    & 8  & 5   & 10    & 9    & 7  & 3    & 6  & {\bf 1}    & 10  \\
\hline
a=3           & 3  & 9    & 5  & 4 & {\bf 1} & 3    & 9  & 5    & 4  & 1     & 5   \\
\hline
a=4           & 4  & 5    & 9  & 3 & {\bf 1} & 4    & 5  & 9    & 3  & 1    & 5 \\
\hline
a=5           & 5  & 3    & 4  & 9 & {\bf 1} & 5    & 3  & 4    & 9  & 1    & 5   \\
\hline
a=6           & 6  & 3    & 7  & 9   & 10    & 5    & 8  & 4    & 2  & {\bf 1}    & 10  \\
\hline
a=7           & 7  & 5    & 2  & 3   & 10    & 4    & 6  & 9    & 8  & {\bf 1}    & 10  \\
\hline
a=8           & 8  & 9    & 6  & 4   & 10    & 3    & 2  & 5    & 7  & {\bf 1}    & 10  \\
\hline
a=9           & 9  & 4    & 3  & 5 & {\bf 1} & 9    & 4  & 3    & 5  & 1    & 5   \\
\hline
a=10         & 10  & {\bf 1}   & 10  & 1   & 10    & 1   & 10  & 1   & 10  & 1    & 2   \\
\hline
\end{tabular}
\end{center}
\caption{Values of $a^i {\rm ~mod~} 11,  1 \leq a,i<11$ and according order of $a$ mod $m$}
\label{SrcArith3a}\label{expmod11}
\end{table}

@@

\begin{table}[ht]
\begin{center}
\begin{tabular}{|l||c|c|c|c|c|c|c|c|c|c|c|c|c|c|c|c|c|c|c|c|c|c|c|c|c|}
\hline
 $a\setminus i$ & 1            & 2            & 3 & 4 & 5 & 6 & 7 & 8 & 9 & 10 & 11 & 12     & $ord_{45}(a)$       & $J(45)$ \\
\hline
\hline                                                       
1             & 1              & 1   & 1   & 1   & 1   & 1   & 1   & 1   & 1    & 1    & 1    & 1 & 1              & 24  \\
\hline
2             & 2              & 4   & 8  & 16  & 32  & 19  & 38  & 31  & 17   & 34   & 23    & 1 & 12             & 24 \\
\hline
3             & 3              & 9  & 27  & 36  & 18   & 9  & 27  & 36  & 18    & 9   & 27   & 36  & ---            & 24 \\
\hline
4             & 4             & 16  & 19  & 31  & 34   & 1   & 4  & 16  & 19   & 31   & 34    & 1  & 6              & 24 \\
\hline
5             & 5             & 25  & 35  & 40  & 20  & 10   & 5  & 25  & 35   & 40   & 20   & 10  & ---            & 24 \\
\hline
6             & 6             & 36  & 36  & 36  & 36  & 36  & 36  & 36  & 36   & 36   & 36   & 36  & ---            & 24 \\
\hline
7             & 7              & 4  & 28  & 16  & 22  & 19  & 43  & 31  & 37   & 34   & 13    & 1  & 12             & 24 \\
\hline
8             & 8             & 19  & 17   & 1   & 8  & 19  & 17   & 1   & 8   & 19   & 17    & 1  & 4              & 24 \\
\hline
9             & 9             & 36   & 9  & 36   & 9  & 36   & 9  & 36   & 9   & 36    & 9   & 36  & ---            & 24 \\
\hline
10           & 10             & 10  & 10  & 10  & 10  & 10  & 10  & 10  & 10   & 10   & 10   & 10  & ---            & 24 \\
\hline
11           & 11             & 31  & 26  & 16  & 41   & 1  & 11  & 31  & 26   & 16   & 41    & 1  & 6              & 24 \\
\hline
12           & 12              & 9  & 18  & 36  & 27   & 9  & 18  & 36  & 27    & 9   & 18   & 36  & ---            & 24 \\
\hline
\end{tabular}
\end{center}
\caption{Values of $a^i{\rm ~mod~}45, 1\leq a,i<13$}
\label{SrcArith3b}\label{expmod45}
\end{table}

@@

\item[{\bf 2.}] Select an arbitrary $e \in \{2, \cdots, n-1\}$ such
                that\footnote{%
                \label{foot:Selection-of-e}%
                It is recommended by cryptanalytic reasons, but not
                necessary to make RSA work, to select $e$ such that: \\
                $\max(p,q)  <  e <  J(n) - 1$.}: \\
                $e$ is relatively prime to $J(n) = (p-1)*(q-1)$. \\
                We can then ``throw away'' $p$ and $q$.\footnote{%
                The procedure also allows us to select $d$ freely and then 
                calculate $e$. However, this has practical disadvantages.
                We usually want to be able to encrypt messages ``quickly'',
                which is why we choose a public exponent $e$ such that 
                it has a short bit length compared to the modulus $n$ and 
                as few binary ones as possible (e.g. $2^{16} + 1$).
                So a fast exponentiation is possible when encrypting.                
                We want to select the publicly known $e$ to be an
                advantageous value that allows the exponential calculation
                to be performed quickly during encryption. The prime numbers
                $3, 17$ and $65537$ have proved to be particularly practical
                for this purpose.
                The most often used number is $65537 = 2^{16}+1$, or in
                binary: 
                $10\cdots 0\cdots 01$ (this number is prime and therefore
                relatively prime to many other numbers).
                }

\item[{\bf 3.}] Select $d \in \{1, \cdots, n-1\}$ with $e*d \equiv 1  
                {\rm ~(mod~} J(n))$, i.e. $d$ is the multiplicative inverse
                of $e$ modulo $J(n)$\footnote{For reasons of security, 
                $d$ should not be too small.}$^,$\footnote{We start by
		determining either $d$ or $e$ depending 
                on the implementation.}. We can then ``throw away'' $J(n)$.
    \begin{itemize}
        \item[] $\rightarrow (n, e)$ is the public key $P$.
        \item[] $\rightarrow (n, d)$ is the private key $S$ (only $d$ must be kept secret).
    \end{itemize}

\item[{\bf 4.}] For encryption, the message represented as a (binary) number is divided
                into parts such that each part of the number is less than $n$.

\item[{\bf 5.}] Encryption of the plaintext (or the parts of it) $M \in \{1, \cdots, n-1\}$:
                $$C = E ( (n, e); M ) := M^e {\rm ~(mod~} n).$$

\item[{\bf 6.}] For decryption, the cipher text represented as a binary number is divided
                into parts such that each part of the number is less than $n$.

\item[{\bf 7.}] Decryption of the cipher text (or the parts of it) $C \in \{1, \cdots, n-1\}$:
                $$M = D ( (n, d); C ) := C^d {\rm ~(mod~} n).$$
\end{itemize}

The numbers $d, e$ and $n$ are usually extremely large (e.~g.\ $d$ and
$e$ $300$ bits, $n$ $600$ bits).\\

@@

\noindent Concretely for the RSA method does this mean:
\vskip - 1em
\begin{enumerate}
  \item It is hard to calculate $J(n)$ for big compounds $n$ and
  \item It is hard to calculate the prime factors of big compounds $n$
        \index{Factorization!factorization problem}.
\end{enumerate}
\vskip +1em

@@

\begin{table}[ht]
\begin{center}
\begin{tabular}{|c|cccc|}
\hline
	& {\bf Decimal digits} & {\bf Binary digits} & {\bf Factored on} & {\bf Factored by} \\
\hline
	C307 & 		307 & 1017 & May, 2007 & Jens Franke et al. \\
	RSA-200 & 	200 & 663 & May, 2005 & Jens Franke et al. \\
	RSA640\footnotemark & 	193 & 600 & November, 2005 & Jens Franke et al. \\
	C176 & 		176 & 583 & May, 2005 & Kazumaro Aoki et al. \\
	RSA576 & 	174 & 576 & December, 2003 & Jens Franke et al. \\
	RSA-160 & 	160 & 530 & April, 2003 & Jens Franke et al. \\
	C158 & 		158 & 523 & January, 2002 & Jens Franke et al. \\
	RSA-155	&	155 & 512 & August, 1999 & Herman te Riele et al. \\
\hline
\end{tabular}
\caption{The current factoring records (as of July 2009)}    % Eyecatcher_neue-Mersenne 
\label{factorizationrecords}
\end{center}
\end{table} 
\footnotetext{%
A research group of the GISA solved this challenge which was awarded with
20,000 US dollar using the GNFS method. The researchers needed about five
months to divide this number into its both 320 bit long prime factors.\\
RSA Labs offers its challenges since the beginning of the 1990th.
The next bigger challenge awarded with 30,000 US dollars is RSA-704.\\
See:
\href{http://www.heise.de/newsticker/meldung/print/65957}
{\texttt{http://www.heise.de/newsticker/meldung/print/65957}}
}

@@

\begin{table}[ht]
\begin{center}
\begin{tabular}{|c|l||c|l|}
\hline
Character & Numerical value & Character & Numerical value \\
\hline
\hline
Blank    & 0   & M  & 13 \\
A        & 1   & N    & 14 \\ 
B        & 2   & O    & 15 \\ 
C        & 3   & P    & 16 \\  
D        & 4   & Q    & 17 \\ 
E        & 5   & R    & 18 \\ 
F        & 6   & S    & 19 \\  
G        & 7   & T    & 20 \\  
H        & 8   & U    & 21 \\ 
I        & 9   & V    & 22 \\   
J       & 10   & W    & 23 \\  
K       & 11   & X    & 24 \\ 
L       & 12   & Y    & 25 \\
&              & Z    & 26 \\
\hline
\end{tabular}
\end{center}
\hypertarget{Grossbuchstaben-Alphabet}{}    
\caption{Capital letters alphabet}\index{Capital letters alphabet}
\label{alphacode}
\end{table}
\vskip +20 pt

@@

% ---------------------------------------------------------------------------
\newpage
\hypertarget{NumberTheory_Sage_Number-of-RSA-keys}{}
\subsection{How many RSA keys exist within a given modulo range?}
\label{l:NumberTheory_Sage_Number-of-RSA-keys}{}

The RSA encryption procedure was described in chapter \ref{RSA} (``\nameref{RSA}'').
Steps 1 to 3 constitute key generation, steps 4 and 5 are the encryption:
\begin{itemize}
  \item[{\bf 1.}] Select two distinct random prime numbers $p$ and $q$
                  and calculate $n = p*q$.\\
                  The value $n$ is called the RSA modulus.

  \item[{\bf 2.}] Select an arbitrary $e \in \{2, \cdots, n-1\}$ such that: \\
                  $e$ is relatively prime to $J(n) = (p-1)*(q-1)$. \\
                  We can then ``throw away'' $p$ and $q$.

  \item[{\bf 3.}] Select $d \in \{1, \cdots, n-1\}$ with $e*d \equiv 1  
                  {\rm ~(mod~} J(n))$, i.e. $d$ is the multiplicative inverse
                  of $e$ modulo $J(n)$. We can then ``throw away'' $J(n)$.
    \begin{itemize}
        \item[] $\rightarrow (n, e)$ is the public key $P$.
        \item[] $\rightarrow (n, d)$ is the private key $S$ (only $d$ must be kept secret).
    \end{itemize}

  \item[{\bf 4.}] For encryption, the message represented as a (binary) number
                  is divided into parts such that each part of 
                  the number is less than $n$.

  \item[{\bf 5.}] Encryption of the plaintext (or the parts of it) $M \in \{1, \cdots, n-1\}$:
                  $$C = E ( (n, e); M ) := M^e {\rm ~(mod~} n).$$
\end{itemize}

The default way to crack a given RSA ciphertext $C$ would be to use the public key
and to try to factorize $n$. Then you can go through the steps 2 and 3 and calculate
the private key $e$, which is normally used to decrypt a ciphertext.

According to the ``prime number theorem''
(described in chapter \ref{thm-pz-pi-x} ``\nameref{thm-pz-pi-x}'')
the number of prime numbers $PI(x)$ is asymptotic to  $x / ln(x)$.
So in a given range $[x,y]$ there are about $ (y / ln(y)) - (x / ln(x))$
primes. If you have a given $n$ then there are about $n / ln(n)$ different possible
values for the prime $p$ (not considering the more specific requirements for selecting
appropriate values for $p$ as described above).
